% Created 2020-10-12 Mon 19:22
% Intended LaTeX compiler: lualatex
\documentclass[11pt]{article}
\usepackage{graphicx}
\usepackage{grffile}
\usepackage{longtable}
\usepackage{wrapfig}
\usepackage{rotating}
\usepackage[normalem]{ulem}
\usepackage{amsmath}
\usepackage{textcomp}
\usepackage{amssymb}
\usepackage{capt-of}
\usepackage{hyperref}
\usepackage{tabularx}
\usepackage{etoolbox}
\makeatletter
\def\dontdofcolorbox{\renewcommand\fcolorbox[4][]{##4}}
\AtBeginEnvironment{minted}{\dontdofcolorbox}
\makeatother
\usepackage[newfloat]{minted}
\author{Mark Armstrong}
\date{October 12th, 2020}
\title{Computer Science 3MI3 – 2020 homework 5\\\medskip
\large “Fizzbuzz”-ing in Ruby}
\hypersetup{
   pdfauthor={Mark Armstrong},
   pdftitle={Computer Science 3MI3 – 2020 homework 5},
   pdfkeywords={},
   pdfsubject={Iterating on a simple programming task in Ruby.},
   pdfcreator={Emacs 27.0.90 (Org mode 9.4)},
   pdflang={English},
   colorlinks,
   linkcolor=blue,
   citecolor=blue,
   urlcolor=blue
   }
\begin{document}

\maketitle
\tableofcontents


\section*{Introduction}
\label{sec:org89f6b8d}
The “fizzbuzz” problem is a very simple programming task,
sometimes used in interviews to check for a basic understanding
of iterating and branching constructs.

We will investigate various possible approaches to
this problem in Ruby, as a way to become comfortable
with the language.
We begin with the familiar looping statements,
and then move to using “higher-order” methods,
as well as solving a generalisation of the problem.

\section*{Boilerplate}
\label{sec:org4d758c0}
\subsection*{Submission procedures}
\label{sec:org1063a8f}
\subsubsection*{Submission method}
\label{sec:orgd39f7e8}

Homework should be submitted to your McMaster CAS Gitlab respository
in the \texttt{cs3mi3-fall2020} project.

Ensure that you have \textbf{pushed} the commits to the remote repository
in time for the deadline, and not just committed to your local copy.

\subsubsection*{Naming requirements}
\label{sec:org284620e}

Place all files for the homework
inside a folder titled \texttt{hn}, where \texttt{n} is the number of the homework.
So, for homework 1, the use the folder \texttt{h1}, for homework 2 the folder \texttt{h2}, etc.
Ensure you do not capitalise the \texttt{h}.

Unless otherwise instructed in the homework questions,
place all of your code for the homework
in a single file in the \texttt{hn} folder named \texttt{hn.ext},
where \texttt{ext} is the appropriate extension for the language used
according to this list:
\begin{itemize}
\item For Scala, \texttt{ext} is \texttt{sc}.
\item For Prolog, \texttt{ext} is \texttt{pl}.
\item For Ruby, \texttt{ext} is \texttt{rb}.
\item For Clojure, \texttt{ext} is \texttt{clg}.
\end{itemize}
If multiple languages are used in the homework,
submit a \texttt{hn.ext} file for each language.

\begin{center}
\textbf{If the language supports multiple different file extensions,}
\textbf{you must still follow the extension conventions above.}
\end{center}

\begin{center}
\textbf{Incorrect naming of files may result in up to a 10\% deduction in your grade.}
\end{center}

\subsubsection*{Do not submit testing or diagnostic code}
\label{sec:org6db9589}

Unless you are instructed to do so in the homework questions,
\textbf{you should not submit testing code with your homework submission}.

This includes
\begin{itemize}
\item any \texttt{main} function,
\item any \texttt{print} statements which output information
\textbf{that is not directly requested as console output}
\textbf{in the homework questions}.
\end{itemize}

If you do not wish to remove diagnostic print statements manually,
you will have to find a way to ensure that they disabled
in your final submission.

For instance, by using a wrapper on the print function or macros.

\subsubsection*{Due date and allowance for technical difficulties}
\label{sec:org5ea5d86}

Homework is due on the second Sunday following its release,
by the end of the day (midnight).
Submissions past 00:00 may not be considered.

If you experience technical difficulties
leading up to the submission time,
please contact Mark \textbf{ASAP} with the details of the problem
and, if possible, attach the current state of your homework
to the communication.
This information will help ensure we are able
to accept your submission once the technical difficulties are resolved.

\subsection*{Proper conduct for coursework}
\label{sec:orgaa7194a}
\subsubsection*{Individual work}
\label{sec:org345efcb}

Unless explicitely stated in the homework questions,
all homework in this course is intended
to be \emph{individually completed}.

You are welcome to discuss the content of the homework in
the public forum of the class Microsoft Teams team homework channel,
though obviously solutions or partial solutions should not
be posted or described.

Private discussions about the homework cannot reasonably be
forbidden, but such discussions should follow the same guidelines
as public discussions.

\begin{center}
\textbf{Inappopriate collaboration via private discussions}
\textbf{which is later discovered by course staff}
\textbf{may be considered academic dishonesty.}

When in doubt, make the discussion private, or report its contents
to the course staff by making a note of it
in your homework.
\end{center}

To clarify what is considered appropriate discussions
of homework content, here are some examples:
\begin{enumerate}
\item Discussing the language features introduced or needed for the homework.
\begin{itemize}
\item Such as relevant builtin datatypes
and datatype definition methods and their general use.
\item Code snippets that are not partial solutions to the homework
are welcome and encouraged.
\end{itemize}
\item Questions of the form “What is meant by \texttt{x}?”,
“Does \texttt{x} really mean \texttt{y}?” or “Is there a mistake with \texttt{x}?”
\begin{itemize}
\item Of course, questions of those form which would be answered
by partial solutions are not considered appropriate.
\end{itemize}
\item Questions or advice about errors that may be encountered.
\begin{itemize}
\item Such as “If you see a \texttt{scala.MatchError} you should
probably add a catch-all \texttt{\_} case to your \texttt{match} expressions.”
\end{itemize}
\end{enumerate}

\subsubsection*{Language library resources}
\label{sec:org3eae2f9}

Unless explicitely stated in the questions,
it is not expected that you will use any language library resources
in the homeworks.

Possible exceptions to this rule include implementations
of datatypes we discuss in this course, such as lists
or options/maybes, if they are included in a standard library
instead of being builtin.

\emph{Basic} operations on such types would also be allowed.
\begin{itemize}
\item For instance, \texttt{head}, \texttt{tail}, \texttt{append}, etc. on lists
would not require explicit permission to be used.
\item More complex operations such as sorting procedures
would require permission before you used them.
\end{itemize}

Additionally, the standard \emph{higher-order} operations
including \texttt{map}, \texttt{reduce}, \texttt{flatten}, and \texttt{filter} are permitted generally,
unless the task is to implement such a higher-order operator.

\section*{Part 1: Fizzbuzzing by loops                            [5 points]}
\label{sec:org554814b}
In Ruby, create a method \texttt{fizzbuzzLooper} which, given a list
(presumably of integers, though it may contain any type)
creates a new list whose elements
are the elements of the original list
converted into strings, unless they are
\begin{itemize}
\item an integer divisible by both \texttt{3} and \texttt{5}, in which case
they are replaced by \texttt{"fizzbuzz"},
\item an integer divisible by \texttt{3} but not by \texttt{5}, in which case
they are replaced by \texttt{"fizz"}, or
\item an integer divisible by \texttt{5} but not by \texttt{3}, in which case
they are replaced by \texttt{"buzz"}.
\end{itemize}

You may want to make use of the \texttt{to\_s} method on integers;
by convention, \texttt{to\_s} on any type converts
objects of that type into strings.

(Technically, your method should probably work
given any type of collection, not just lists;
but the result should be a list in any case.)

Your \texttt{fizzbuzzLooper} must make use of some manner of
looping construct.
\begin{itemize}
\item Such as a \texttt{loop}, \texttt{while} loop or \texttt{for} loop.
\end{itemize}

\begin{center}
\textbf{Because this is a fairly trivial programming task,}
\textbf{the marking of this question}
\textbf{(and to some extent the marking of the remainder of the homework)
*will take \emph{elegance} more into account than usual.}
(meaning you are expected to follow good coding practices,
especially \emph{not repeating yourself}.)
\end{center}

\section*{Part 2: Fizzbuzzing by iterators (higher-order methods) [10 points]}
\label{sec:orga390521}
Construct another method \texttt{fizzbuzzIterator},
whose behaviour is identical to \texttt{fizzbuzzLooper},
but which is defined using an “iterator” method
rather than a looping construct.

See this online
\href{https://ruby-doc.com/docs/ProgrammingRuby/html/tut\_containers.html}{tutorial}
on collections and iterators.
In particular, look into the iterators \texttt{each} and \texttt{collect}.

These iterators take a \emph{block} as argument.
A block is, essentially, Ruby's “lambda” construct.
Blocks are delimited by braces, \texttt{\{\}}, and may have
arguments, which are listed at the beginning and
delimited by pipes, \texttt{||}.
So the anonymous function \texttt{λ x → x + 1} would
be written \texttt{\{ |x| x + 1 \}} in Ruby.

So for instance,
\begin{minted}[breaklines=true]{ruby}
[1,2,3].each { |x| puts(x) }
\end{minted}

outputs each element of the list \texttt{[1,2,3]}.

\section*{Part 3: Generalising fizzbuzzing                        [20 points]}
\label{sec:orgc1c3169}
We now consider a slight generalisation to
the fizzbuzzing problem, which we will call “zuzzing”.

To generalise the problem, we assume that we may have
several rules which should be applied to the elements of this list,
instead of just the two (if it's disible by 3, output “fizz”,
if it's divisible by 5, output “buzz”.)

We want to create a method which accomodates any number of rules,
and where the rules can be arbitrary predicates
on the elements of the list
(not just “\texttt{λ x → x is divisible by \textasciitilde{}k}”.)

To represent this multitude of rules, we use a list of lists,
assuming each of the lists within the list contain two elements;
\begin{itemize}
\item the first element being a \texttt{lambda} for the rule, and
\item the second element being a \texttt{lambda} for the string
associated with that rule.
\begin{itemize}
\item We use a \texttt{lambda} here as well so that the resulting string
may depend upon the element.
\end{itemize}
\end{itemize}

(The keyword \texttt{lambda} applied to a block allows you to store
that block using a variable, or in our case, in a list;
we are still essentially using blocks in this question.)

For instance, to get the original behaviour of “fizzbuzz”
using this “zuzz”, we would use the rules
\begin{minted}[breaklines=true]{ruby}
rules = [[lambda { |x| x % 3 == 0 }, lambda { |x| "fizz" }],
         [lambda { |x| x % 5 == 0 }, lambda { |x| "buzz" }]]
\end{minted}
as in
\begin{minted}[breaklines=true]{ruby}
zuzzer([1,2,3,4,5,6,7,8,9,10,11,12,13,14,15],rules)
\end{minted}

The reason we use a list of lists of lambdas here to encode the rules,
rather than a hash table or other construct,
is that \emph{the order of the rules matters}.
If more than one rule applies to an element, all such rules
should be applied \emph{in order} to build the resulting string.
For instance, with the “fizzbuzz” rules above, notice
that the “fizz” rule comes before the “buzz” rule
so that if an element is divisible by both 3 and 5,
the result is \texttt{"fizzbuzz"}, not \texttt{"fizz"}, \texttt{"buzz"} or \texttt{"buzzfizz"}.

Create the method \texttt{zuzzer}.

\section*{Part 4: Generalised fizzbuzzing in Scala                [10 bonus points]}
\label{sec:orgea96343}
Implement the generalised fizzbuzzing operation from part 3 in Scala.

Make what you feel are necessary adjustments to the types or
implementation details, and describe your choices in comments.
Your solution may be (sometimes subjectively) judged based on
the choices you make. The purpose of the comments is then
for you to convince us your choices are appropriate.

\section*{Part 5: Generalised fizzbuzzing in Prolog               [10 bonus points]}
\label{sec:org3bd3958}
Implement the generalised fizzbuzzing operation from part 3 in Scala.

Make what you feel are necessary adjustments to the types or
implementation details, and describe your choices in comments.
Your solution may be (sometimes subjectively) judged based on
the choices you make. The purpose of the comments is then
for you to convince us your choices are appropriate.

\section*{Testing}
\label{sec:org099aee9}
Unit tests for the requested methods/functions
are available \href{./testing/h5/h5t.rb}{here}.
The contents of the unit test file are also repeated below.

The tests can be run by placing the \texttt{h4t.sc} file
in the same directory as your \texttt{h4.sc} file, and
running the following command.
\begin{minted}[breaklines=true]{shell}
ruby h5t.rb
\end{minted}
You may also use \texttt{irb h5t.rb}, which will echo the script
as it is run.

If you wish to add more tests yourself,
see the \href{https://en.wikibooks.org/wiki/Ruby\_Programming/Unit\_testing}{documentation}
for Ruby.

\subsection*{Automated testing via Docker}
\label{sec:org6728845}
The Docker setup and usage scripts are available at the following links.
Their contents are also repeated below.
\begin{itemize}
\item \href{./testing/h5/Dockerfile}{Dockerfile}
\item \href{./testing/h5/docker-compose.yml}{docker-compose.yml}
\item \href{./testing/h5/setup.sh}{setup.sh}
\item \href{./testing/h5/run.sh}{run.sh}
\end{itemize}
Place them into your \texttt{h5} directory where your \texttt{h5.rb} file
and the \texttt{h5t.rb} (linked to above) file exist,
then run \texttt{setup.sh} and \texttt{run.sh}.

You may also refer to the README
for this testing setup and those files
\href{https://github.com/armkeh/principles-of-programming-languages/tree/master/homework/testing/h5}{on the course GitHub repo}.

Note that the use of the \texttt{setup.sh} and \texttt{run.sh} scripts assumes
that you are in a \texttt{bash} like shell; if you are on Windows,
and not using WSL or WSL2, you may have
to run the commands contained in those scripts manually.

\subsection*{The tests}
\label{sec:orgef0df95}
The contents of the testing script are repeated here.

\href{./testing/h5/h5t.rb}{h5t.rb}
\begin{minted}[breaklines=true]{ruby}
require_relative "h5"
require "test/unit"

# BEGIN constants for use in testing ---------------------------
# We use global variables, which is not recommended in Ruby,
# but which makes it simpler to share these values here.
$upto15 = [0,1,2,3,4,5,6,7,8,9,10,11,12,13,14,15]

$upto15_fizzbuzzed = ["fizzbuzz","1","2","fizz","4","buzz",
                     "fizz","7","8","fizz","buzz",
                     "11","fizz","13","14","fizzbuzz"]

$fizzbuzz_rules = [[lambda { |x| x % 3 == 0 }, lambda { |x| "fizz" }],
                   [lambda { |x| x % 5 == 0 }, lambda { |x| "buzz" }]]
# END constants for use in testing -----------------------------

# BEGIN fizzbuzzLooper tests -----------------------------------
class TestLooper < Test::Unit::TestCase

  def test_simple
    assert_equal($upto15_fizzbuzzed,fizzbuzzLooper($upto15))
  end

end
# END fizzbuzzLooper tests -------------------------------------

# BEGIN fizzbuzzLooper tests -----------------------------------
class TestIterator < Test::Unit::TestCase

  def test_simple
    assert_equal($upto15_fizzbuzzed,fizzbuzzIterator($upto15))
  end

end
# END fizzbuzzLooper tests -------------------------------------

# BEGIN fizzbuzzLooper tests -----------------------------------
class TestZuzzer < Test::Unit::TestCase

  def test_simple
    assert_equal($upto15_fizzbuzzed,zuzzer($upto15,$fizzbuzz_rules))
  end

end
# END fizzbuzzLooper tests -------------------------------------
\end{minted}

\subsection*{The Docker setup}
\label{sec:org7351740}
The contents of the Docker setup files and scripts are repeated here.

\href{./testing/h5/Dockerfile}{Dockerfile}
\begin{minted}[breaklines=true]{docker}
FROM ruby:2.7.2-buster

# Set the name of the maintainers
MAINTAINER Habib Ghaffari Hadigheh, Mark Armstrong <ghaffh1@mcmaster.ca, armstmp@mcmaster.ca>

# Set the working directory
WORKDIR /opt/h5
\end{minted}

\href{./testing/h5/docker-compose.yml}{docker-compose.yml}
\begin{minted}[breaklines=true]{yaml}
version: '2'
services:
  service:
    build: .
    image: 3mi3_h5_docker_image
    volumes:
      - .:/opt/h5
    container_name: 3mi3_h5_container
    command: ['ruby', 'h5t.rb']
\end{minted}

\href{./testing/h5/setup.sh}{setup.sh}
\begin{minted}[breaklines=true]{shell}
docker-compose build --force-rm
\end{minted}

\href{./testing/h5/run.sh}{run.sh}
\begin{minted}[breaklines=true]{shell}
# Run the container
docker-compose up --force-recreate
# Stop the container after finishing the test run
docker-compose stop -t 1
\end{minted}

\href{./testing/h5/README.md}{README.md}
\begin{minted}[breaklines=true]{text}
# Instructions for automated testing using Docker

We have already created a `Dockerfile` here which specifies
all the necessary packages, etc., for compiling and running your code.
You only need to follow the instructions below to see 
the results of unit tests designed to check your implementation.

## Setup
We use `docker-compose` and its configuration file to build the image.
Assuming you have `docker` and `docker-compose` installed,
simply execute
```shell script
./setup.sh
```
to generate the image.

## Prepare your code for the running the tests
You only need to place the `h5t.rb` unit test file and
the `run.sh` file in the same directory as your `h5.rb` source file.

## Running the tests
As with the build process, we have already put
the configuration needed for running the test inside `docker-compose.yml`.
Simply execute
```shell script
./run.sh
```
to run and see the results of the tests.
\end{minted}
\end{document}
