% Created 2020-11-04 Wed 11:30
% Intended LaTeX compiler: lualatex
\documentclass[11pt]{article}
\usepackage{graphicx}
\usepackage{grffile}
\usepackage{longtable}
\usepackage{wrapfig}
\usepackage{rotating}
\usepackage[normalem]{ulem}
\usepackage{amsmath}
\usepackage{textcomp}
\usepackage{amssymb}
\usepackage{capt-of}
\usepackage{hyperref}
\usepackage{tabularx}
\usepackage{etoolbox}
\makeatletter
\def\dontdofcolorbox{\renewcommand\fcolorbox[4][]{##4}}
\AtBeginEnvironment{minted}{\dontdofcolorbox}
\makeatother
\usepackage[newfloat]{minted}
\usepackage{amsthm}
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\usepackage{unicode-math}
\usepackage{unicode}
\author{Mark Armstrong}
\date{Fall 2020}
\title{Types\\\medskip
\large Principles of Programming Languages}
\hypersetup{
   pdfauthor={Mark Armstrong},
   pdftitle={Types},
   pdfkeywords={},
   pdfsubject={Introduction to types},
   pdfcreator={Emacs 27.0.91 (Org mode 9.4)},
   pdflang={English},
   colorlinks,
   linkcolor=blue,
   citecolor=blue,
   urlcolor=blue
   }
\begin{document}

\maketitle

\section{Preamble}
\label{sec:org4ea34bc}

\subsection{Notable references}
\label{sec:orgd1e8925}

\begin{itemize}
\item Robert W. Sebesta, “Concepts of Programming Languages” (10th edition)
\begin{itemize}
\item Chapter 6, Data Types
\end{itemize}

\item Benjamin Pierce,
“\href{https://ebookcentral.proquest.com/lib/mcmu/detail.action?docID=3338823}{Types and Programming Languages}”
\begin{itemize}
\item Chapter 11, Simple Extensions
\begin{itemize}
\item Unit, Tuples, Sums, Variants, Lists.
\end{itemize}
\item Chapter 13, References
\begin{itemize}
\item Reference types.
\end{itemize}
\item Chapter 20, Recursive types
\end{itemize}
\end{itemize}

\subsection{{\bfseries\sffamily TODO} Table of contents}
\label{sec:org7b6e1ad}

\begin{scriptsize}
\begin{itemize}
\item \hyperref[sec:org4ea34bc]{Preamble}
\end{itemize}
\end{scriptsize}

\section{Introduction}
\label{sec:org50d4033}

This section introduces the concepts of \emph{types},
a particularly useful language safety feature.

Common simple types and methods of building new types are discussed,
as well as some more advanced topics.

The discussion in this section is fairly “encyclopedic”;
the next section on the typed λ-calculus give some insight about
how some of these types can be incorporated into a programming system.

\section{Properties of type systems}
\label{sec:org17a5557}

In the previous notes, we have discussed
\begin{itemize}
\item polymorphism and
\item static/dynamic typing
\end{itemize}
which are two important properties of a type system.

Here we discuss some other commonly discussed properties,
before discussing in the following sections
what is arguably the most important property:
what types a language might have.

\subsection{“Strong” and “weak” typing}
\label{sec:orgf7ae74c}

These are comparative terms.
\begin{itemize}
\item We'll consider them a subjective criteria.
\end{itemize}

“Strongly typed”
\begin{itemize}
\item Languages are frequently called strongly typed.
\begin{itemize}
\item But less frequently do they state what they mean by that.
\item The term is used inconsistently.
\begin{itemize}
\item “C is a strongly typed, weakly checked language”
– Dennis Ritchie, creator of C
\end{itemize}
\end{itemize}
\end{itemize}

“Weakly typed” simply means not strongly typed.

\subsubsection{So what does “strongly typed” mean?}
\label{sec:orgae2d5b6}

We will take it to mean “type clashes are restricted”.
\begin{itemize}
\item A \emph{type clash} being an instance where one type is expected
but a different, incompatible type is provided.
\begin{itemize}
\item Such as added a string to an integer.
\end{itemize}
\item That definition does not make a good objective property.
\begin{itemize}
\item What does restricted mean?
\begin{itemize}
\item Is it a warning or an error?
\item Does type casting violate this?
\end{itemize}
\item What qualifies as a type clash?
\begin{itemize}
\item Is implicit type casting allowed?
\end{itemize}
\end{itemize}
\end{itemize}

\subsubsection{Exercise: Examples of implicit type casting}
\label{sec:org4cc82f1}

What are the results of these two Javascript output statements?
\begin{minted}[breaklines=true]{js}
console.log(1 + 2 + "3")
console.log("1" + 2 + 3)
\end{minted}

How about the following C++ code?
(This is arguably not quite an example of implicit type casting.)
\begin{minted}[breaklines=true]{c++}
#include<iostream>

int main() {
  std::cout << ("Hello" + 1);
}
\end{minted}

\subsection{Explicit and implicit typing}
\label{sec:org3546a06}

Languages may require annotations on variables and functions
(\emph{explicit typing}) or allow them to be omitted (\emph{implicit typing}).
\begin{itemize}
\item Implicit typing does not weaken the typing system in any way!
\begin{itemize}
\item A very common misconception.
\end{itemize}
\item In general, type inference is an undecidable problem
(its not guaranteed that the compiler/interpreter can
determine the type).
\begin{itemize}
\item Most languages have relatively simple type systems,
and this is not a problem.
\item We will likely later study the \emph{Hindley-Milner} type system
for the λ-calculus, which is used for many
functional languages and supports parametric polymorphism.
\end{itemize}
\end{itemize}

\subsubsection{Implicit typing by name}
\label{sec:orgc00dfce}

Some languages make type annotations a part of the name,
or annotate names with sigils to indicate type details.
\begin{itemize}
\item In older versions of Fortran, names beginning with
\texttt{i}, \texttt{j} or \texttt{k} were for integer variables,
and all variables were of floating point.
\item In Perl, names beginning with the sigil
\begin{itemize}
\item \texttt{\$} have scalar type,
\item \texttt{@} have array type,
\item \texttt{\%} have hash type, and
\item \texttt{\&} have subroutine type.
\end{itemize}
\end{itemize}

\section{Atomic types}
\label{sec:org9098990}

We begin our discussion of what types languages have
with what are usually the “simplest” types: \emph{atomic} types.
\begin{itemize}
\item Atomic in the sense that they cannot be broken down any further.
\item Sometimes called \emph{primitive} or \emph{basic}.
\end{itemize}

\subsection{Ubiquitous basic types}
\label{sec:orgfaa19f5}

Most languages have at least these atomic types.
\begin{itemize}
\item \textbf{Integers}; \texttt{int}
\begin{itemize}
\item Including possibly signed, unsigned, short, and/or long variants.
\end{itemize}
\item \textbf{Floating point} numbers
\begin{itemize}
\item Including possibly single precision and double precision variants.
\end{itemize}
\item \textbf{Characters}
\begin{itemize}
\item Sometimes an alternate name for the byte type (8-bit integers).
\end{itemize}
\item \textbf{Booleans}
\begin{itemize}
\item Which, of course, can be stored in a bit,
but will usually be stored in at least a word (4 bits)
or perhaps even a byte (8 bits)
for convenience of memory access.
\end{itemize}
\end{itemize}

\subsection{Singleton and empty types}
\label{sec:orga3d1a18}

\begin{itemize}
\item \textbf{Unit} (the \emph{singleton} type)
\begin{itemize}
\item Sometimes called \texttt{void}, \texttt{nil}-type, \texttt{null}-type or \texttt{none}-type.
\begin{itemize}
\item In C-like languages, you cannot store something of type \texttt{void}.
\begin{itemize}
\item But this value still implicitely exists;
a function of type \texttt{void} can certainly
return, implying it returns something of type \texttt{void}.
\end{itemize}
\item Commonly represented as the type of 0-ary tuples,
whose only element is \texttt{()}.
\end{itemize}
\end{itemize}
\item \textbf{Empty}
\begin{itemize}
\item Unlike a singleton type, which has a single value
(called \texttt{nil}, \texttt{null} or \texttt{none}), there is
(or should be) \textbf{nothing} in the empty type.
\begin{itemize}
\item This might be violated by using error values.
\item For instance, in Haskell, the term referred to as \texttt{bottom} or ⊥
—which refers to any computation which never completes—
 is a member of all types, even the empty type.
\end{itemize}
\item No function with return type \texttt{Empty} should
ever be able return.
\end{itemize}
\end{itemize}

\subsubsection{Singleton and empty types in Haskell}
\label{sec:org3490bab}

\begin{minted}[breaklines=true]{haskell}
data Unit = Unit -- This type Unit has a single constructor, also called Unit.
data Empty -- Empty has no constructors.

-- Previous versions of Haskell required a workaround to write Empty,
-- as a constructor list was mandatory;
-- this alternate definition has a constructor also called EmptyAlt,
-- but using this constructor requires an argument of type EmptyAlt
-- to already exist, so should never actually be usable.
data EmptyAlt = EmptyAlt EmptyAlt


-- The bottom can be defined by unending recursion.
bottom = bottom
  
x = Unit

-- We can cheat and get Empty/EmptyAlt "values" by using bottom.
y = bottom :: Empty
z = bottom :: EmptyAlt

-- Check the types of these values
:t x   -- reports x :: Unit
:t y   -- reports y :: Empty
:t z   -- reports z :: EmptyAlt
\end{minted}

\subsection{Implementation of atomic types}
\label{sec:orgb18e59f}

When we discussed the pure untyped λ-calculus,
we discussed the process of \emph{encoding} the integers and booleans
as functions, since they were not included in the language.
\begin{itemize}
\item We also mentioned that we can add constants for them
to the language, forming an \emph{unpure} untyped λ-calculus.
\end{itemize}

This raises a question we can ask about
“practical” programming languages as well;
\begin{itemize}
\item are the “atomic” (“primitive”, “basic”) types \emph{truly} atomic
(primitive, basic), or are they represented
using one of the language's abstractions?
\item We have discussed the fact that in Scala and Ruby, which we call
“purely object-oriented”, even these atomic types are classes!
\begin{itemize}
\item Whereas in Java and C++, they are not;
there, they are “primitives” which exist
outside the object-oriented abstraction.
\end{itemize}
\end{itemize}

\subsection{Atom or symbol types}
\label{sec:org1f22ea5}

Many languages include a type of \emph{atoms} or \emph{symbols},
which are essentially interned strings.
\begin{itemize}
\item Strings which are immutable,
and of which there is only one copy in memory.
\end{itemize}

Specifically, decendents of Lisp and Prolog tend to have a symbol type.

For instance, in Ruby:
\begin{minted}[breaklines=true]{ruby}
# A symbol in Ruby begins with a :
x = :hello

# We can intern strings using an intern method
y = "hello".intern

# The equal? method checks if two values are the same object.
# These all evaluate to true, since there's only one copy of the symbol hello.
puts x.equal?(y)
puts :hello.equal?(x)
puts y.equal?(:hello)

# But this to false, because these are two copies of the same string!
puts "hello".equal?("hello")
\end{minted}

\subsection{Ordinal types}
\label{sec:org183d5b7}

Many languages include a means of defining other \emph{finite} types.
Instances include
\begin{itemize}
\item enumeration types (\texttt{enum}'s) and
\item subset/subrange types.
\end{itemize}

For instance, Pascal supports both enumerations
and subranges.
(This example based on one from the \href{https://wiki.freepascal.org/Data\_type}{Free Pascal and Lazarus Wiki}.)
\begin{minted}[breaklines=true]{pascal}
type
  // An enumeration type; unless specified,
  // the first element is implicitely assigned the ordinal value 0,
  // the second ordinal value 1, etc.
  DaysOfWeek = (Sunday, Monday, Tuesday, Wednesday,
                Thursday, Friday, Saturday);

  // A subrange type.
  // In this case, DaysOfWorkWeek contains the ordinals 1,2,3,4 and 5.
  DaysOfWorkWeek = Monday..Friday;
\end{minted}

\subsection{Less common numeric types}
\label{sec:orgdc13bbe}

\begin{itemize}
\item \textbf{Complex} numbers
\begin{itemize}
\item Especially for scientific computation.
\end{itemize}
\item \textbf{Decimal} (representation of) numbers
\begin{itemize}
\item Especially for business (monetary) applications.
\item There are decimal numbers that cannot be properly represented
using binary (e.g. \texttt{0.3 = 0.010011}, repeating)
\item Not included in all languages because
they cannot be efficiently represented.
\begin{itemize}
\item For instance, to store a decimal digit directly
(without conversion to binary) takes at least 4 bits.
\item There are 10 possibilities, too many for 3 bits (which can
only have 8 different states.)
\item But 4 bits could represent 16 states (6 more than needed.)
\item For memory access reasons, some such implementations
even use 8 bits (one byte) per decimal digit.
\end{itemize}
\end{itemize}
\end{itemize}

\subsubsection{Complex numbers in C\#}
\label{sec:org7e7d963}

(This example code taken from the
\href{https://docs.microsoft.com/en-us/dotnet/api/system.numerics.complex?view=netcore-3.1}{.NET documentation}.)
\begin{minted}[breaklines=true]{csh}
using System;
using System.Numerics;

public class Example
{
   public static void Main()
   {
      // Create a complex number by calling its class constructor.
      Complex c1 = new Complex(12, 6);
      Console.WriteLine(c1);

      // Assign a Double to a complex number.
      Complex c2 = 3.14;
      Console.WriteLine(c2);

      // Cast a Decimal to a complex number.
      Complex c3 = (Complex) 12.3m;
      Console.WriteLine(c3);

      // Assign the return value of a method to a Complex variable.
      Complex c4 = Complex.Pow(Complex.One, -1);
      Console.WriteLine(c4);

      // Assign the value returned by an operator to a Complex variable.
      Complex c5 = Complex.One + Complex.One;
      Console.WriteLine(c5);

      // Instantiate a complex number from its polar coordinates.
      Complex c6 = Complex.FromPolarCoordinates(10, .524);
      Console.WriteLine(c6);
   }
}
// The example displays the following output:
//       (12, 6)
//       (3.14, 0)
//       (12.3, 0)
//       (1, 0)
//       (2, 0)
//       (8.65824721882145, 5.00347430269914)
\end{minted}

\subsubsection{Decimal numbers in C\#}
\label{sec:org7f8112f}

(This example code taken from the
\href{https://docs.microsoft.com/en-us/dotnet/api/system.decimal?view=netcore-3.1}{.NET documentation}.)
\begin{minted}[breaklines=true]{csh}
// Keeping my fortune in Decimals to avoid the round-off errors.
class PiggyBank {
    protected decimal MyFortune;

    public void AddPenny() {
        MyFortune = Decimal.Add(MyFortune, .01m);
    }

    public decimal Capacity {
        get {
            return Decimal.MaxValue;
        }
    }

    public decimal Dollars {
        get {
            return Decimal.Floor(MyFortune);
        }
    }

    public decimal Cents {
        get {
            return Decimal.Subtract(MyFortune, Decimal.Floor(MyFortune));
        }
    }

    public override string ToString() {
        return MyFortune.ToString("C")+" in piggy bank";
    }
}
\end{minted}

\section{Structured data}
\label{sec:org5181265}

Of course, we rarely want to deal only with “atomic” data.
More commonly, we are interested in \emph{collections} of data,
or \emph{alternatives} between different data types;
these are captured in the notion of \emph{structured} data.

Specifically, we consider
\begin{itemize}
\item products,
\item sequences,
\begin{itemize}
\item including arrays and lists
\end{itemize}
\item “sets”
\begin{itemize}
\item (really associative arrays (maps/hashes/tables))
\end{itemize}
\item and unions/variants.
\end{itemize}

\subsection{Homogeneous or heterogeneous}
\label{sec:org4870800}

An important design decision for any structured type
is whether it is \emph{homogeneous} or \emph{heterogeneous}.
\begin{itemize}
\item “Heterogeneous” structures store elements of differing types.
\item “Homogeneous” store only elements of the same type.
\end{itemize}

\subsection{Product types (tuples)}
\label{sec:org07ee43d}

A \emph{heterogeneous} collection of a \emph{fixed} number of elements.
\begin{itemize}
\item Implemented by, for instance,
\begin{itemize}
\item \texttt{struct}'s or records,
\begin{itemize}
\item both of which have labelled fields,
\end{itemize}
\item tuples,
\begin{itemize}
\item which are often be implemented as records with
specially named fields, such as “\texttt{first}” or “\texttt{\_1}”, and
\end{itemize}
\item classes,
\begin{itemize}
\item which have \emph{methods} as well as fields.
\end{itemize}
\end{itemize}
\item In lower level languages, programmers may be concerned
with the alignment/packing of the data.
\begin{itemize}
\item See \href{http://www.catb.org/esr/structure-packing/}{this overview} of structure packing in C,
and the sort of memory inefficiency that can result
from poorly ordering the fields of a \texttt{struct}.
\end{itemize}
\end{itemize}

\subsection{Array types}
\label{sec:org4635eca}

Arrays are an abstraction of finite sequences of
adjacent memory cells.
\begin{itemize}
\item Programmers are guaranteed (or required to work around) certain properties.
\begin{itemize}
\item \texttt{O(1)} access/update time for any element.
\begin{itemize}
\item Implying elements are stored in adjacent memory cells
(a \emph{contiguous block} of memory cells.)
\item Also implying each element is of a known, usually constant, size.
\begin{itemize}
\item Heterogeneous arrays can store \emph{references} to elements of differing types
instead of elements of those types themselve to maintain this property.
\end{itemize}
\item To perform a lookup, just look \texttt{element\_space * index} bits
past the start of the array.
\end{itemize}
\item \texttt{O(n)} insert time (if inserting is possible.)
\begin{itemize}
\item The (possibly \texttt{n}-many) elements after the insertion point
must be shifted.
\end{itemize}
\item It may be computationally costly or impossible to modify length.
\end{itemize}
\end{itemize}

We can classify arrays by where and how their memory is allocated.

\subsection{Classification of arrays based on memory allocation: data segment and stack}
\label{sec:orgd2c2b01}

\begin{itemize}
\item Static arrays
\begin{itemize}
\item (Memory) allocation is static, and presumably in the data segment.
\item Subscript ranges are statically bound.
\end{itemize}
\item Fixed stack-dynamic arrays
\begin{itemize}
\item Allocation is dynamic and on the stack,
done when the declaration of the array is reached.
\item Subscript ranges are statically bound.
\end{itemize}
\item Stack-dynamic arrays
\begin{itemize}
\item Allocation is dynamic and on the stack,
done when the declaration of the array is reached.
\item Subscript ranges are dynamically bound when the declaration is reached.
\item After declaration, subscript range and storage remain fixed.
\end{itemize}
\end{itemize}

\subsection{Classification of arrays based on memory allocation: heap}
\label{sec:org620fc2f}

\begin{itemize}
\item Fixed heap-dynamic arrays
\begin{itemize}
\item Allocation is dynamic, done when the user program requests.
\item Subscript ranges are dynamically bound at allocation time.
After allocation, subscript range and storage remain fixed.
\end{itemize}
\item Heap-dynamic arrays
\begin{itemize}
\item Allocation is dynamic, done when the user program requests.
\item Subscript ranges are dynamically bound at allocation time.
\item Subscript range and storage can change throughout runtime!
\item These are commonly implemented as \emph{array-lists}, discussed shortly.
\end{itemize}
\end{itemize}

\subsection{List types}
\label{sec:org39f3a2e}

Lists are simply an abstract notion of sequences.
\begin{itemize}
\item May be implemented by arrays or by structures such as linked lists.
\item Often we do not \texttt{O(1)} access time for lists.
\item But we do have better flexibility;
\begin{itemize}
\item appending or prepending an element into a linked list
only requires changing a pointer,
\item and inserting into a list only requires changing \texttt{O(n)} pointers.
\end{itemize}
\end{itemize}

Persistance of data is more feasible using linked lists;
parts of the list can be reused.
\begin{itemize}
\item For instance, two immutable lists \texttt{[1,2,3,4]} \texttt{[5,6,3,4]} can share
the storage for the \texttt{3} and \texttt{4} elements.
\end{itemize}

Lazily (non-strictly) constructed lists may even be “infinite”.
For instance, the infinite list of 1's in Haskell:
\begin{minted}[breaklines=true]{haskell}
ones = 1 :: ones
\end{minted}

\subsection{Array lists}
\label{sec:orgc14bfb5}

An \emph{array list} type provides the flexibility of a list,
but the performance of an array (except for the occasional reallocation.)
\begin{itemize}
\item Array lists are stored in a continguous block of memory cells.
\item And the block is just reallocated when the array grows too large.
\begin{itemize}
\item Reallocating memory is costly,
and it requires copying the elements to the new memory,
and so there is a performance dip when this is needed.
\item Generally, \emph{twice as much} memory as is needed is allocated.
\begin{itemize}
\item So as the size of the list grows, so does the extra memory allocated
at each reallocation.
\item This mitigates the cost of reallocation; it becomes less frequent
as time goes on.
\end{itemize}
\end{itemize}
\end{itemize}

\subsection{“Sets”}
\label{sec:org6391735}

It is notoriously difficult to represent unordered collections
such as sets and bags on computers.
\begin{itemize}
\item Computers are extremely ordered machines; how do we store
unordered collections in ordered memory?
\item When available, “set types” are usually implemented
using \emph{trees} or \emph{associative arrays} (discussed next.)
\end{itemize}

\subsection{Associative array (map, hash, table) types}
\label{sec:org3b2a863}

Associative arrays, also called \emph{hashes}, \emph{maps} or sometimes \emph{tables},
are sets of key/value pairs.
\begin{itemize}
\item Abstracts away the ordering of the sequence.
\begin{itemize}
\item (Though we could order the keys, and so impose
an order on the collection.)
\end{itemize}
\item The programmer can imagine they are lists of key/value pairs.
\begin{itemize}
\item And they may even be implemented that way.
\begin{itemize}
\item But in practical languages, they are more often
implemented by sorting elements into “buckets”
by a hashing function on the keys.
\item The implementations get quite complex.
\end{itemize}
\end{itemize}
\end{itemize}

\subsection{Unions, variants}
\label{sec:orgc684de5}

Whereas an element of a product type contains
\begin{itemize}
\item a collection of elements of some types,
\end{itemize}
a \emph{union} or variant type contains
\begin{itemize}
\item one element of a selection of types.
\end{itemize}

Unions can be \emph{tagged} or \emph{untagged}.
\begin{itemize}
\item With an untagged union, the runtime does not
keep track of the underlying type of the element.
\item Whereas a tagged union uses a \emph{tag} on the value to identify
the underlying type of the element.
\end{itemize}

Note that union types are unnecessary
in dynamically type checked language.
\begin{itemize}
\item We can think of every variable/argument/procedure's type
as being a union of all possible types.
\end{itemize}

\subsection{Untagged unions}
\label{sec:org9b57c02}

As we've said, with an untagged union,
the runtime does not keep track of the underlying type of elements.
\begin{itemize}
\item So it's type is dynamic! (Amongst the types involved in the union.)
\item Languages that provide untagged unions (today, mainly C and C++)
do not even dynamically check the type.
\item Accessing it as the wrong type simply treats the bits
as if they were of that type; it is not a cast!
\item This is \emph{very} unsafe; it allows for type clashes.
\end{itemize}

\subsubsection{An example usage of untagged unions in C++}
\label{sec:org213c0c3}

We can see the danger of untagged unions in a short C++ example;
we are legally allowed to interpret the bits of an integer
as if they were the bits of a floating point,
which has a very different arrangement in memory!
\begin{minted}[breaklines=true]{c++}
#include<iostream>

union foo {
  int a;
  float b;
};

int main() {
  foo x;

  x.a = 1; // Set x as an integer.
  
  std::cout << x.b; // Treat x as a float, even though it's an int right now.
                    // The typechecker does not complain!
  
  // Outputs 1.4013e-45 during my testing;
  // the result of interpreting the bits of x as a float.
}
\end{minted}

For a more practical usage of untagged unions, see the example
implementation of a language of integer expressions in C++,
provided in homework 6.

\subsection{Tagged unions and pattern matching}
\label{sec:orgd7ab20b}

Tagged unions are also known as
\begin{itemize}
\item \emph{sum} and \emph{either} types, or
\item as \emph{variant} types when the labels are chosen by the programmer.
\end{itemize}

The introduction of tags or labels makes \emph{pattern matching} a viable
and useful control structure.

Given a variant type such as
\begin{minted}[breaklines=true]{haskell}
data Foo = A Int | B Float
\end{minted}

We can \emph{match} on the tag, using placeholder variables
for the values of the underlying types.
\begin{minted}[breaklines=true]{haskell}
bar :: Foo -> Int
bar (A i) = i   -- i is a placeholder for the integer value.
bar (B _) = 0   -- An example of a nameless placeholder, _.
\end{minted}
Many languages also allow the use of constants
instead of placeholder variables, to create more specific cases.

\subsection{Recursive unions}
\label{sec:orga3a93b6}

If a union type is allowed to be \emph{recursive} (meaning that
one of the underlying types of the union is allowed to be
the type being defined as the union),
then union and product types together allow for the definition
of \emph{algebraic datatypes} (discussed later.)

A well-known example is to define lists as a recursive union
of products.
\begin{minted}[breaklines=true]{haskell}
data List = Cons (Int, List) | Empty
\end{minted}

(Using a type parameter and currying, we get this more flexible
definition more in the style of most functional languages.)
\begin{minted}[breaklines=true]{haskell}
data List A = Cons A (List A) | Empty
\end{minted}

\section{References}
\label{sec:orgefb2c55}

Pointer and reference types capture the notion of a
memory address.
\begin{itemize}
\item Not just alternate namings! They have very different properties.
\item Pointers are a numerical representation, and can be manipulated
as numbers to access \emph{adjacent} memory locations.
\begin{itemize}
\item Pointer arithmetic.
\item Note: adding \texttt{1} to a pointer does not shift by \texttt{1} bit;
it shifts by the size of the type being pointed to.
\end{itemize}
\item A reference is a more abstract notion;
it cannot (or at least should not) be manipulated as a pointer can.
\end{itemize}

\subsection{Referencing/dereferencing}
\label{sec:orgafb2a41}

In a language with reference types, we typically have
\begin{itemize}
\item a \emph{dereference} operator to access the value
stored at that reference
\begin{itemize}
\item (written \texttt{*} in C and many C-like languages, and
\item written \texttt{!} in ML and languages descended from it),
\end{itemize}
\end{itemize}
and we often have
\begin{itemize}
\item a \emph{reference} operator to obtain a reference to a stored value
\begin{itemize}
\item (written \texttt{\&} in C and many C-like languages.)
\end{itemize}
\end{itemize}

\subsection{Garbage collection and dangling references}
\label{sec:orgcb725e3}

When discussing memory binding, we previously discussed
the concept of \emph{garbage collection} and mentioned
another related problem:
\begin{itemize}
\item \emph{dangling} or \emph{wild} references.
\end{itemize}

A dangling or wild reference is a reference value
which refers to a memory location that has already been deallocated.
\begin{itemize}
\item The contents of deallocated memory are usually not specified
by the language, so accessing it can result in undefined behaviour.
\item This \emph{should} only occur if the language allows explicit deallocation
of memory (instead of only using implicit deallocation
through garbage collection.)
\end{itemize}

There are mechanisms to prevent against the use of dangling references.

\subsection{Preventing the use of dangling references; tombstones}
\label{sec:org13804b1}

With the \emph{tombstone} approach, every reference is in fact
a reference to a cell of memory called a \emph{tombstone},
which then refers to the value being stored in memory.
\begin{itemize}
\item So there is an additional level of indirection.
\item When memory is deallocated, the tombstone is left,
but its reference is set to a null pointer.
\begin{itemize}
\item So it can no longer be used; accessing a “nulled” tombstone
will result in an error.
\end{itemize}
\item This is costly both in time (two dereferences are needed for each access)
and in space (the tombstone must remain indefinitely.)
\end{itemize}

\subsection{Preventing the use of dangling references; locks and keys}
\label{sec:org604f9db}

With the \emph{lock-and-keys} approach, a reference value is in fact
a pair of a special \emph{key} value and the reference itself.
\begin{itemize}
\item And referenced memory cells have allocated with them a
special \emph{lock} value which matches those keys.
\item When memory is freed, the lock value for that cell is wiped.
\item So the keys of any remaining references are no longer valid.
\item This is also costly in time
\begin{itemize}
\item (a comparison must be made between
the lock and the key at each access)
\end{itemize}
\item and in space
\begin{itemize}
\item (the lock and key values take space on every reference
and every memory cell)
\end{itemize}
\item but not so much as tombstones.
\end{itemize}

\section{Further advanced topics}
\label{sec:orgc1a1fe9}

Depending upon time at the end of the course,
we may return to discuss more about types.
\end{document}
